---
title: "R Assignments"
author: "Prof. Dr. Christoph Richard, Leonie Wicht, Anna Vandebosch and Theresa Schmid"
date: "2024-01-04"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R1. Runs test of randomness 

The data sequence R1.dat describes the amount of dead fish in kg per year, which has been found along a certain river within a period of 16 years between 1995 and 2010. Use the runs test in order to test the hypothesis that the amount of dead fish obeys random fluctuations.

1.  Read the data sequence into your computer and determine the scale of measure.
2.  Dichotomise the data sequence using the algorithm discussed in the lecture.
3.  Use the command `rle` in order to determine the number of runs. What is the test decision? Give all details of the calculation.
4.  What do you learn concerning the validity of the randomness hypothesis?

*Remark:* Read the documentation of the command `rle`. Using `$lengths` you can extract the vector of lengths from the output of `rle`.


## R1. Solution

1. Read the data sequence into your computer and determine the scale of measure.
```{r}
data <- scan(file="Data/R1dat.sec")
data <- as.vector(data)

```


The variable (amount of dead fish in kg per year, which has been found along a certain river within a period of 16 years between 1995 and 2010) has a ratio scale, it is metric (continuous).


2. Dichotomise the data sequence using the algorithm discussed in the lecture.

We will use the median as a cut-off criterion. A data point is converted to 0 if it is equal or less than the median and 1 otherwise (in R FALSE corresponds to 0 and TRUE to 1).


```{r}
(data_dich <- data > median(data))
```

3. Use the command `rle` in order to determine the number of runs. What is the test decision? Give all details of the calculation.

The command rle (Run Length Encoding) computes the lengths and values of runs of equal values in a vector (cf R help). We can confirm the lengths of the runs comparing it to the output of the previous part.

```{r}
runs <- rle(data_dich)
(rl <- runs$lengths)
```

The number of runs here is 6.

The mean number of runs in a sequence of length n, n0 zeros and n1 ones is


$$
\mu = \frac{2n_0n_1}{n} + 1 \stackrel{\text{here}}{=} 2 \cdot \frac{8 \cdot 8}{16} + 1 = 9
$$


The variance of the number of runs is

$$
\sigma^2 = \frac{(\mu-1)(\mu -2)}{(n-1)} \stackrel{\text{here}}{=} \frac{8 \cdot 7}{15} = 3.733
$$

We accept the hypothesis of randomness if the number of runs is between $\mu - 2\sigma$ and $\mu + 2\sigma$.

$$
[\mu - 2\sigma, \mu + 2\sigma] = [5.13, 12.86]
$$

As the number of runs in our data set (6) is included in the interval, we have to accept the hypothesis that the underlying process of finding dead fish is random.
 
4. What do you learn concerning the validity of the randomness hypothesis?

The approximation that is used in this test (95% of random experiments have number of runs in the above interval) is only good for n0, n1 ≥ 20. Thus we can’t say anything specific about the randomness of the data set because it is to small. Besides, the runs test has low sensitivity (high error of type II; runs test is test against randomness), thus we might accept randomness although it is false.


## R2. Data analysis

There were altogether 308 soccer games during the 2008 european soccer championship qualification. The number of goals per match is stored in the file R2.dat. 

1. Create a table of the absolute goal frequencies.
2. Create a bar chart of the relative goal frecuencies. Create a histogram of relative goal frecuencies. Compare the graphics. Can you identify the "outlier"?
3. Draw the empirical cumulative distribution function.

## R2. Solution

1. Create a table of the absolute goal frequencies.

```{r}

data <- scan(file = "Data/R2dat.sec")
data <- as.vector(data)
(table <- table(data))

```

2. Create a bar chart of the relative goal frecuencies. Create a histogram of relative goal frecuencies. Compare the graphics. Can you identify the "outlier"?

```{r}

barplot(table / length(data), xlab = "Goals",
        ylab = "Relative frequency", main = "Relative goal frequencies")

hist(data, freq = FALSE, xlab = "Goals",
     ylab = "Relative frequency", main = "Relative goal frequencies")

```

The differences are:

- the bar chart has one bar for every number of goals that ocurred in these 308 matches, so you don’t have space between the outlier and the rest
- the histogram has the breaks in the natural numbers so you can’t see easily the frequency of a specific number of goals, but here you see that there was an outlier without looking on the x-axis


3. Draw the empirical cumulative distribution function.

```{r}

plot(ecdf(data), xlab = "Goals", ylab = "Cumulative frequency",
     main = "Empirical cumulative distribution function")

```
